Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-08-21 manuscript/plos/review_1.tex
added computational resources, explicit how Pearson correlation is computed, and first grammarly pass

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable, producing different results when the input data changes slightly. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, encoding an ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. An elegant explanation has been proposed with the omnigenic model: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated, so that many other genes are able to alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we characterize six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. Our pipeline can easily be applied to other diseases.

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

@@

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}). It selected 100 SNPs, both from all aforementioned regions and new ones (\nameref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation. This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparably (poor) classification performance to L1-penalized logistic regression (Fig~\ref{fig:benchmark}B), while providing more interpretable solutions.

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (BH adjusted P\=/value < 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}B). Another prominent difference is their size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI failed to recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ, as measured by the median centrality (Table~\ref{tab:gene_solutions}) and the number of connected components (Fig~\ref{fig:solution_overview}D). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences in the underlying mechanisms. It linked different parts of SigMod's solution network to four processes: protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~<~0.03). Interestingly, dmGWAS solution is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~<~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~<~0.04).

@@

Despite their differences, there are additional common themes. All obtained solution subnetworks have lower association P\=/values than the whole PPIN (median P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}A). This exemplifies the trade-off between statistical significance and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes. Additionally, the solution subnetworks presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value < 0.03). Second, the genes in four solution subnetworks displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value < 1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (Supplementary table~\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non associated SNPs.

@@

A pathway enrichment analysis of the genes in the consensus network also shows similar pathways as the individual solutions. We found two involved mechanisms: \emph{mitochondrial translation} and \emph{attenuation phase}. The former is supported by genes like \emph{MRPS30} (VEGAS2 P\=/value~=~0.001), which encode a mitochondrial ribosomal protein and was also linked to breast cancer susceptibility \cite{quigley_5p12_2014}. Interestingly, increased mitochondrial translation has been found in cancer cells \cite{Yu2016Repositioning}, and its inhibition proposed as a therapeutic target. With regards to attenuation phase of heat shock response, it involves three Hsp70 chaperones: HSPA1A, HSPA1B, and HSPA1L. The genes encoding these proteins are all near each other at 6p21, in the region known as HLA. In fact, out of the 22 SNPs that map to any of these three genes, 9 map to all of them, and 4 to two, making hard to disentangle their effects. \emph{HSPA1A} was the most strongly associated gene (VEGAS2 P\=/value~=~8.37~\texttimes{}~10\textsuperscript{-4}).

Topologically the consensus consists of a connected component composed of 49 genes, and multiple smaller subnetworks. Among the latter, 19 genes are in subnetworks containing a single gene or two connected nodes, implying that they do not have a consistently altered neighborhood, but are strongly associated themselves and hence picked by at least two methods. The opposite would be the case of highly central genes in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}D). This suggests that they were selected because they were on the shortest path between two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus network is COPS5, a component of the COP9 signalosome which regulates multiple signalling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or BCAC (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus subnetwork have consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

@@

We compared the results of different network methods to the European sample of the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017}, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a family history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network metods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC, in an equivalent way to the one described in Section~\ref{methods:node_score}. The solutions provided by the different network approaches overlap significantly with BCAC findings (Fisher's exact test P\=/value \textless~0.019). The gene-based network methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (Fig~\ref{fig:solution_overview}C). Interestingly, while SConES GI at the SNP-level achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

@@

We compared the six network methods in a 5-fold subsampling setting (Section \ref{methods:comparison}). Specifically, we measured five properties (Fig~\ref{fig:benchmark}): size of the solution subnetwork; sensitivity and specificity of an L1-penalized logistic regression classifier on the selected SNPs; stability; and computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs. The largest solutions came from SConES GI (6\,256.6 SNPs), and dmGWAS (4\,255.0 SNPs). To determine whether the selected SNPs could be used for patient classification, we computed the performance of the classifier on the \emph{test dataset} (Fig~\ref{fig:benchmark}B). The different classifiers displayed similarly poor sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming only at minimizing prediction error. It should be considered that, although these performances are low, we do not expect to separate cases from controls well using exclusively genetic data.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\linewidth]{./figures/figure_4.pdf}
  \caption{{ \bf Comparison of network-based GWAS methods on GENESIS. } Each method was run 5 times on a random subset containing 80\% of the samples, and tested on the remaining samples (Section \ref{methods:comparison}). As LEAN did not select any gene, it was excluded from all panels except \textbf{D}. \textbf{(A)} Number of SNPs selected by each method and number of SNPs in the active set found by the classifier. Points are the average over the 5 runs; lines represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison. For reference, the active set of Lasso using all the SNPs included, on average, 154\,117.4 SNPs. \textbf{(B)} Sensitivity and specificity on test set of the L1-penalized logistic regression trained on the features selected by each of the methods. In addition, the performance of the classifier trained on all SNPs is displayed. Points are the average over the 5 runs; lines represent the standard error of the mean. \textbf{(C)} Pairwise Pearson correlations of the solutions used by different methods. A Pearson correlation of 1 means the two solutions are the same. A Pearson correlation of 0 means that there is no SNP in common between the two solutions. \textbf{(D)} Runtime of the evaluated methods, by type of network used (gene or SNP). For gene network-based methods, inverted triangles represent the runtime of the algorithm itself, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics.}
  \label{fig:benchmark}
  \end{figure}

Another desirable quality of an algorithm is the stability of the solution with regards to small changes in the input (Section~\ref{methods:algorithm_comparison}). Both heinz and LEAN displayed a high stability in our benchmark, consistently selecting the same genes and no genes over the 5 subsamples, respectively (Fig~\ref{fig:benchmark}C). Conversely, the other methods displayed similarly low stabilities.

@@

As shown above, and despite their similarities, the network methods produced remarkably different solutions. This is due to the particularities of each methods, and directly or indirectly provide information about the dataset. For instance, the fact that LEAN did not return any biomarker implies that there is no gene such that both itself and its environment are on average strongly associated with the disease.

@@

On the other end of the spectrum, dmGWAS, HotNet2, and SigMod produced large solutions. dmGWAS' subnetwork is the least associated subnetwork on average. This is due to the greedy framework it uses, which has a bias for larger solutions \cite{nikolayeva_network_2018}. It considered all nodes at distance 2 of the examined subnetwork, and accepted a weakly associated genes if it was linked to another, strongly associated one. This is exacerbated when the results of successive greedy searches are aggregated, leading to a large, tightly connected cluster of unassociated genes (Fig~\ref{fig:issues}A). This relatively low signal-to-noise ratio combined with the large solution requires additional analyses to draw conclusions, such as enrichment analyses. In the same line, HotNet2's subnetwork is even harder to interpret, being composed of 440 genes divided into 135 subnetworks. Lastly, SigMod misses some of the most strongly associated, breast cancer susceptibility genes in the dataset, like \emph{FGFR2} and \emph{TOX3}.

@@

Most of the network methods, including the consensus, were highly unstable, raising questions about the reliability of the results. We built a new, stable consensus network using the genes selected most often across the 30 solutions obtained by running the 6 methods on 5 different splits of the data (Section \ref{methods:comparison}). Such a network is expected to capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 of the solutions, which corresponded to 1\% of all genes selected at least once. The resulting stablility-based consensus was composed of 68 genes (Fig \ref{fig:stable-consensus}). This network shares most of the proterties of the consensus: breast cancer susceptibility genes are overrepresented (P\=/value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P\=/values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively); the selected genes are more central than average (P\=/value = 1.1 \texttimes{} 10\textsuperscript{-14}); and a considerable number of nodes (19) are isolated.

@@

In addition to a weight, nodes have other properties, provided by the topology of the graph. In this article we focus on two of those: degree centrality, and betweenness centrality. The degree centrality, or degree, is the number of edges that a node has. The betweenness centrality, or betweenness, is the number of times a node participates in the shortest paths between two other nodes.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to the size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable, producing different results when the input data changes slightly. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, encoding a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely a lack of statistical power and interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. An elegant explanation has been proposed with the omnigenic model: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated so that many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we characterize six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. Our pipeline can easily be applied to other diseases.

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as it is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

@@

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}). It selected 100 SNPs, both from all aforementioned regions and new ones (\nameref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the protein-protein interaction network (PPIN, Section~\ref{methods:networks}). In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation. This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparably (poor) classification performance to L1-penalized logistic regression (Fig~\ref{fig:benchmark}B), while providing more interpretable solutions.

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value < 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus. These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}B). Another prominent difference is their size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contained only 4 genes. While SConES GI failed to recover any protein coding gene, by dealing with SNP networks it retrieved four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}). Their topologies also differ, as measured by the median centrality (Table~\ref{tab:gene_solutions}) and the number of connected components (Fig~\ref{fig:solution_overview}D). Only two methods have more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have less than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}. Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences in the underlying mechanisms. It linked different parts of SigMod's solution network to four processes: protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~<~0.03). Interestingly, the dmGWAS solution is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). But, additionally, it includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~<~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~<~0.04).

@@

Despite their differences, there are additional common themes. All obtained solution subnetworks have lower association P\=/values than the whole PPIN (median P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}A). This exemplifies the trade-off between statistical significance and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes. Additionally, the solution subnetworks presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value < 0.03). Second, the genes in four solution subnetworks displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value < 1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (Supplementary table~\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non-associated SNPs.

@@

A pathway enrichment analysis of the genes in the consensus network also shows similar pathways as the individual solutions. We found two involved mechanisms: \emph{mitochondrial translation} and \emph{attenuation phase}. The former is supported by genes like \emph{MRPS30} (VEGAS2 P\=/value~=~0.001), which encode a mitochondrial ribosomal protein and was also linked to breast cancer susceptibility \cite{quigley_5p12_2014}. Interestingly, increased mitochondrial translation has been found in cancer cells \cite{Yu2016Repositioning}, and its inhibition proposed as a therapeutic target. With regards to the attenuation phase of heat shock response, it involves three Hsp70 chaperones: HSPA1A, HSPA1B, and HSPA1L. The genes encoding these proteins are all near each other at 6p21, in the region known as HLA. In fact, out of the 22 SNPs that map to any of these three genes, 9 map to all of them, and 4 to two, making it hard to disentangle their effects. \emph{HSPA1A} was the most strongly associated gene (VEGAS2 P\=/value~=~8.37~\texttimes{}~10\textsuperscript{-4}).

Topologically the consensus consists of a connected component composed of 49 genes, and multiple smaller subnetworks. Among the latter, 19 genes are in subnetworks containing a single gene or two connected nodes, implying that they do not have a consistently altered neighborhood, but are strongly associated themselves and hence picked by at least two methods. The opposite would be the case of highly central genes in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}D). This suggests that they were selected because they were on the shortest path between two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus network is COPS5, a component of the COP9 signalosome which regulates multiple signaling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or BCAC (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus subnetwork have consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

@@

We compared the results of different network methods to the European sample of the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017}, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a family history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network methods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC, in an equivalent way to the one described in Section~\ref{methods:node_score}. The solutions provided by the different network approaches overlap significantly with BCAC findings (Fisher's exact test P\=/value \textless~0.019). The gene-based network methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (Fig~\ref{fig:solution_overview}C). Interestingly, while SConES GI at the SNP-level achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

@@

We compared the six network methods in a 5-fold subsampling setting (Section \ref{methods:benchmark}). Specifically, we measured five properties (Fig~\ref{fig:benchmark}): size of the solution subnetwork; sensitivity and specificity of an L1-penalized logistic regression classifier on the selected SNPs; stability; and computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs. The largest solutions came from SConES GI (6\,256.6 SNPs), and dmGWAS (4\,255.0 SNPs). To determine whether the selected SNPs could be used for patient classification, we computed the performance of the classifier on the \emph{test dataset} (Fig~\ref{fig:benchmark}B). The different classifiers displayed similarly poor sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming only at minimizing prediction error. It should be considered that, although these performances are low, we do not expect to separate cases from controls well using exclusively genetic data.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\linewidth]{./figures/figure_4.pdf}
  \caption{{ \bf Comparison of network-based GWAS methods on GENESIS. } Each method was run 5 times on a random subset containing 80\% of the samples, and tested on the remaining samples (Section \ref{methods:benchmark}). As LEAN did not select any gene, it was excluded from all panels except \textbf{D}. \textbf{(A)} Number of SNPs selected by each method and number of SNPs in the active set found by the classifier. Points are the average over the 5 runs; lines represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison. For reference, the active set of Lasso using all the SNPs included, on average, 154\,117.4 SNPs. \textbf{(B)} Sensitivity and specificity on the test set of the L1-penalized logistic regression trained on the features selected by each of the methods. The performance of the classifier trained on all SNPs is also displayed. Points are the average over the 5 runs; lines represent the standard error of the mean. \textbf{(C)} Pairwise Pearson correlations of the solutions used by different methods. A Pearson correlation of 1 means the two solutions are the same. A Pearson correlation of 0 means that there is no SNP in common between the two solutions. \textbf{(D)} Runtime of the evaluated methods, by type of network used (gene or SNP). For gene network-based methods, inverted triangles represent the runtime of the algorithm itself, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics.}
  \label{fig:benchmark}
  \end{figure}

Another desirable quality of an algorithm is the stability of the solution with regards to small changes in the input (Section~\ref{methods:algorithm_comparison}). Both heinz and LEAN displayed high stability in our benchmark, consistently selecting the same genes and no genes over the 5 subsamples, respectively (Fig~\ref{fig:benchmark}C). Conversely, the other methods displayed similarly low stabilities.

@@

As shown above, and despite their similarities, the network methods produced remarkably different solutions. This is due to the particularities of each method, and directly or indirectly provide information about the dataset. For instance, the fact that LEAN did not return any biomarker implies that there is no gene such that both itself and its environment are on average strongly associated with the disease.

@@

On the other end of the spectrum, dmGWAS, HotNet2, and SigMod produced large solutions. dmGWAS' subnetwork is the least associated subnetwork on average. This is due to the greedy framework it uses, which has a bias for larger solutions \cite{nikolayeva_network_2018}. It considered all nodes at distance 2 of the examined subnetwork, and accepted a weakly associated gene if it was linked to another, strongly associated one. This is exacerbated when the results of successive greedy searches are aggregated, leading to a large, tightly connected cluster of unassociated genes (Fig~\ref{fig:issues}A). This relatively low signal-to-noise ratio combined with the large solution requires additional analyses to draw conclusions, such as enrichment analyses. In the same line, HotNet2's subnetwork is even harder to interpret, being composed of 440 genes divided into 135 subnetworks. Lastly, SigMod misses some of the most strongly associated, breast cancer susceptibility genes in the dataset, like \emph{FGFR2} and \emph{TOX3}.

@@

Most of the network methods, including the consensus, were highly unstable, raising questions about the reliability of the results. We built a new, stable consensus network using the genes selected most often across the 30 solutions obtained by running the 6 methods on 5 different splits of the data (Section \ref{methods:benchmark}). Such a network is expected to capture the subnetworks more often found altered, and hence should be more resistant to noise. We used only genes selected in at least 7 of the solutions, which corresponded to 1\% of all genes selected at least once. The resulting stability-based consensus was composed of 68 genes (Fig \ref{fig:stable-consensus}). This network shares most of the properties of the consensus: breast cancer susceptibility genes are overrepresented (P\=/value = 3 \texttimes{} 10\textsuperscript{-4}), as well as genes involved in mitochondrial translation and the attenuation phase (adjusted P\=/values 0.001 and 3 \texttimes{} 10\textsuperscript{-5} respectively); the selected genes are more central than average (P\=/value = 1.1 \texttimes{} 10\textsuperscript{-14}); and a considerable number of nodes (19) are isolated.

@@

In addition to weights, nodes have other properties, provided by the topology of the graph. We focus on two of those: degree centrality, and betweenness centrality. The degree centrality, or degree, is the number of edges that a node has. The betweenness centrality, or betweenness, is the number of times a node participates in the shortest paths between two other nodes.

We also use two matrices that describe two different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_i$ and $v_j$. The \emph{adjacency matrix} $W_G$ contains a 1 when the corresponding nodes are connected, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_G$ is a diagonal matrix which contains the degree of the different nodes.
